\documentclass[11pt]{article}

\usepackage[
% fancytheorems, 
% fancyproofs,
noindent, 
nokoma
%spacingfix, 
]{adam}

% \usepackage{ulem}
% \DeclareRobustCommand{\hsout}[1]{\texorpdfstring{\sout{#1}}{#1}}

\usepackage{cancel}

% \usepackage{titling}
% \setlength{\droptitle}{-4em}


\title{\vspace{-3\baselineskip}\ \\ Examples \& Counterexamples in Group Theory}
% \author{Adam Kelly, Vishal Gupta, Conor Rajan,\\ Conor Doherty, Joe Devine, Hasan Baig}
\author{Christ's Mathmos}
\date{\today}

\begin{document}

\maketitle

This is a collection of groups that have specific interesting properties, such as being a counterexample to some non-obvious statements about groups. Most of these are well known.

\textbf{A group isomorphic to every non-trivial subgroup}. $\Z$.

\textbf{A group of order $n$ which has no subgroup of $k$, where $k \mid n$}. $A_5$, since it is simple and thus has no subgroup of index 2. Another example is $A_4$ (which is the smallest such group), but it's a little harder to show that it works.

\textbf{Two non-isomorphic groups with the same order type}. $C_4 \times C_4$ and $C_2 \times Q_8$. Both have 1 element of order 1, 3 elements of order 2, and 12 elements of order 4. If this property holds for two abelian groups, then they are isomorphic.

\textbf{A non-abelian group with all non-identity elements of order $p$}. The group of upper triangular $3 \times 3$ matrices over $\Z/p\Z$ with 1s on the diagonal.

\textbf{A non-abelian group of order $p^3$}. Same as above.

\textbf{An infinite group whose proper subgroups are all finite}. The group $\left\{k/2^{n} : k, n \in \mathbb{N}, k<2^{n}\right\}$ with addition modulo 1. This is the Prüfer $2$-group.

\textbf{A group $G$ with $N \normal G$ and $H \normal N$ such that $H \not \normal G$}. In $D_8$, we have $\langle s \rangle \normal \langle r^2, s \rangle \normal \langle r, s \rangle = D_4$, but $\langle s \rangle \not \normal D_4$.

\textbf{A group in which every group generated by $n$ elements has a surjective homomorphism to it}. The free group with a basis of $n$ elements.


\textbf{An infinite group with every non-identity element of order 2}. $C_2 \times C_2 \times C_2 \times \cdots$.

\textbf{An infinite non-abelian group with every element of finite order}. $S_3 \times C_2 \times C_2 \times \cdots$.

\textbf{A group $G$ with $G \cong G \times G$}. $G = C_2 \times C_2 \times C_2 \times \cdots$. We can also get $G \cong G \times G \times G \times \cdots$ in the natural way.

\textbf{Two non-zero elements of $\R$ that generate a subgroup not isomorphic to $\Z$}. $1$ and $\sqrt{2}$.

\textbf{A group with two subgroups whose product is not a subgroup}. Consider $D_3$, with the distinct reflections $s, s' \in D_3$. Then $\{e, s\} \times \{e, s'\} \not \leq D_3$.

\textbf{A quotient group of a finite group that is not isomorphic to a subgroup}. $Q_8 / \{-1, 1\}$.

\textbf{A group isomorphic to its automorphism group}. $S_3 \cong \operatorname{Aut}(S_3)$.

\textbf{A non-abelian infinite group where the set of elements of finite order is a subgroup}. $\GL_2(\Q)$. We note that this also holds for all abelian groups, and indeed any group where the elements of finite order commute with each other.

\textbf{A group that is not a semi-direct product}. $Q_8$.

\textbf{A group $G$ with two isomorphic subgroups $H, K$ where $G/H$ and $G/K$ are not isomorphic}. Take $G = C_4 \times C_2$, with $H = \langle (2, 0) \rangle$ and $K = \langle (0, 1)\rangle$. Then $G/H \cong C_2 \times C_2$, and $G/K \cong C_4$.


\end{document}

% Citations: 

% This page for counterexample 
% https://math.stackexchange.com/questions/2333666/group-is-isomorphic-to-direct-product-of-its-subgroups
