\documentclass[a4paper]{article}

\usepackage[
    % fancytheorems, 
    % fancyproofs, 
    % noindent, 
    nokoma
]{adam}

% \documentclass[a4paper]{scrartcl}

% \usepackage[
%     fancytheorems, 
%     fancyproofs, 
%     noindent, 
% ]{adam}

\title{Complex Analysis}
\author{Adam Kelly (\texttt{ak2316@cam.ac.uk})}
\date{\today}

\allowdisplaybreaks

\begin{document}

\maketitle

% This is a short description of the course. It should give a little flavour of what the course is about, and what will be roughly covered in the notes.

This article constitutes my notes for the `Complex Analysis' course, held in Lent 2022 at Cambridge. These notes are \emph{not a transcription of the lectures}, and differ significantly in quite a few areas. 
Still, all lectured material should be covered.
%In 

\section{Analytic Functions}

\subsection{Complex Differentiability}

This course is about complex valued functions of a (single) complex variable, that is, functions
$$
f: A \rightarrow \C, \quad \text{where } A \subset \C.
$$
In this course we will be particularly interested in functions that are \emph{complex differentiable}.

\begin{definition}
    Let $f: U \rightarrow \C$ where $U \subseteq \C$ is open. We say that $f$ is \vocab{differentiable} at $z \in U$ if the limit
    $$
    f'(z) = \lim_{h \rightarrow 0} \frac{f(z + h) - f(z)}{h}
    $$
    exists.
\end{definition}


\begin{itemize}
    \item \emph{Analytic functions}. Complex differentiation and Cauchy-Riemann equations. Examples. Conformal mappings. Informal discussion of branch points, examples of $\log z$ and $z^c$.
    \item \emph{Contour Integration and Cauchy's theorem}. Contour integration (for piecewise continuously differentiable curves). Statement and proof of Cauchy's theorem for star domains. Cauchy's integral formula. Maximum modulus theorem. Louiville's theorem. Fundamental theorem of algebra. Morera's theorem.
    \item \emph{Expansions and singularities}. Uniform convergence of analytic functions. Local uniform convergence. Differentiability of power series. Taylor and Laurent expansions. Principle of isolated zeros. Residue at an isolated singularity. Classification of isolated singularities.
    \item \emph{Residue Theorem}. Winding numbers. Residue theorem. Jordan's lemma. Evaluation of definite integrals by contour integration. Rouche's theorem, principle of the argument. Open mapping theorem.
\end{itemize}

\end{document}
